---
title: "Reducing Fragmentation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reducing Fragmentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
  )
```

## Introduction

In many spatial planning problems, we are not only interested in meeting conservation targets at minimum cost, but also in **how fragmented** the selected areas are. Fragmented solutions can be difficult to manage, may fail to support viable populations, and are often harder to communicate to stakeholders.

This vignette explores different ways to **reduce fragmentation** in a typical prioritizr / spatialplanr workflow, using the Tasmania planning unit dataset as a case study. We will compare four approaches:

**1. Manual boundary penalties in *prioritizr***

**2. Automatic calibration of boundary penalties using `calibrate_cohon_penalty()`**

**3. Minimum patch size rules** using the *minpatch* package

**4. Boundary penalties inside MinPatch** during the whittling stage

Throughout, we will compare these approaches in terms of:

-   **Cost** and total selected area

-   **Fragmentation metrics** (total boundary length, number of patches, patch size)

-   **Runtime**

Our guiding question is:

> **Given the same conservation problem, how do different fragmentation approaches change the cost, pattern, and runtime of our solutions?**

## Data and baseline problem

In this vignette, we use the Tasmania planning unit and feature data from the **prioritizr** package. These data represent a real-world conservation planning problem and are also used in other *prioritizr* examples, which makes it easier to compare workflows.

```{r load-packages, message=FALSE, warning=FALSE}
library(spatialplanr)
library(prioritizr)
library(prioritizrdata)
library(minpatch)

library(dplyr)
library(sf)
library(ggplot2)
library(ggrepel)
library(tibble)
library(tidyr)
library(patchwork)

# install helper if needed
#if (!requireNamespace("remotes", quietly = TRUE)) {
#  install.packages("remotes")
#}

# install rcbc from GitHub
#remotes::install_github("dirkschumacher/rcbc")
#library(rcbc)
```

### Load the Tasmania data

```{r load-tasmania, fig.width=8, fig.height=6}
# load data
tas_pu <- get_tas_pu()  
 
# At present minpatch works with sf objects. Here we convert the data to sf.
tas_features <- get_tas_features() %>% 
  stars::st_as_stars() %>% 
  sf::st_as_sf()

tas <- sf::st_interpolate_aw(tas_features, tas_pu, extensive = FALSE, keep_NA = FALSE, na.rm = FALSE) %>% 
  st_join(tas_pu, join = st_equals)

summary(tas$cost)

features = tas %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(-all_of(c("id", "cost", "locked_in", "locked_out"))) %>% 
  names()

# Convert data to binary again
tas <- tas %>% 
  mutate(across(all_of(features), ~ if_else(.x > 0, 1, 0)))


#To ease runtime: Crop the study area to a manageable size 

bbox <- sf::st_bbox(tas)

frac <- 0.50  # use to adjust the scope of the planning region

bbox_small <- bbox
bbox_small["xmax"] <- bbox["xmin"] + frac * (bbox["xmax"] - bbox["xmin"])
bbox_small["ymax"] <- bbox["ymin"] + frac * (bbox["ymax"] - bbox["ymin"])

tas_small <- sf::st_crop(tas, bbox_small)

tas <- tas_small  

#nrow(tas)
# plot planning units
ggplot(tas) + 
  geom_sf() +
  theme_minimal() +
  labs(
    x = NULL, y = NULL)
```

For this vignette, we will:

-   treat the `cost` column in `tas_pu` as the planning unit cost, and
-   aim to represent 30% of each feature's distribution

Notes:
-   running the whole dataset at 100% causes Cohon calibration to be severely long
-   it is worth lookin into it again (at 100%) using a more powerful machine 

### Build a baseline prioritizr problem

We first define a **baseline problem** with no explicit fragmentation control:

-   **Objective**: minimum set (minimise total cost)
-   **Targets**: 30% relative targets for all features
-   **Decisions**: binary (each planning unit is either selected or not)
-   **Solver**: default solver with a 10% optimality gap

```{r base-problem}

p <- problem(tas, features = features, cost_column = "cost") %>%
  add_min_set_objective() %>%
  add_relative_targets(0.30) %>%  # 30% of each feature
  add_binary_decisions() %>%
  add_cbc_solver(verbose = TRUE)

print(p)

```

### Solve the baseline problem

```{r}
set.seed(1) # for reproducibility

base_time <- system.time({
base_solution <- solve(p)
})

base_time
```

### Plot the baseline solution

```{r base-solution, fig.width=8, fig.height=6}
plot_prioritizr(base_solution)
```

In the next sections, we will use this baseline solution to compute fragmentation metrics (e.g., number and size of patches, total boundary length), and then compare it against solutions obtained using boundary penalties and MinPatch.

## Helper: fragmentation metrics

To compare different fragmentation controls, we will use a small helper function to compute summary metrics for each solution. This lets us build a single table of results across all methods.

```{r metrics-helper}
summarise_solution <- function(label,
                              problem,        # still kept for consistency, but not used now
                              solution_sf,
                              solve_time_sec,
                              min_patch_size = NULL,   # e.g. from MinPatch (m²)
                              boundary_penalty = 0) {  # penalty used in optimisation

  # 1) check & subset selected PUs
  if (!"solution_1" %in% names(solution_sf)) {
    stop("summarise_solution(): column 'solution_1' not found in solution_sf")
  }

  sel <- solution_sf[solution_sf$solution_1 == 1, ]
  n_selected <- nrow(sel)

  # if nothing is selected, return mostly zeros/NA
  if (n_selected == 0) {
    return(tibble(
      method             = label,
      n_selected         = 0L,
      total_area_km2     = 0,
      n_patches          = 0L,
      valid_patches      = NA_integer_,
      median_patch_km2   = NA_real_,
      pu_cost_total      = 0,
      pu_cost_mean       = NA_real_,
      boundary_length_km = 0,
      boundary_cost      = 0,
      total_cost         = 0,
      solver_time_sec    = solve_time_sec
    ))
  }

  # 2) area + patch metrics 

  # area of each selected PU (m²)
  pu_areas_m2    <- as.numeric(sf::st_area(sel))
  total_area_km2 <- sum(pu_areas_m2) / 1e6

  # patches = contiguous blobs of selected PUs
  patches <- sel |>
    sf::st_union() |>
    sf::st_cast("POLYGON")

  patch_areas_m2   <- as.numeric(sf::st_area(patches))
  n_patches        <- length(patch_areas_m2)
  median_patch_km2 <- stats::median(patch_areas_m2) / 1e6

  # valid patches if a min patch size is provided
  if (!is.null(min_patch_size)) {
    min_patch_size_m2 <- as.numeric(min_patch_size)
    valid_patches <- sum(patch_areas_m2 >= min_patch_size_m2)
  } else {
    valid_patches <- NA_integer_
  }

  # boundary length (outer boundary of patches, in km)
  patch_boundaries   <- sf::st_cast(sf::st_boundary(patches), "MULTILINESTRING")
  boundary_length_km <- sum(as.numeric(sf::st_length(patch_boundaries))) / 1000

  # 3) cost metrics ---------------------------------------

  if (!"cost" %in% names(solution_sf)) {
    stop("summarise_solution(): column 'cost' not found in solution_sf")
  }

  pu_cost_total <- sum(sel$cost, na.rm = TRUE)
  pu_cost_mean  <- mean(sel$cost, na.rm = TRUE)

  # boundary cost only matters if a penalty was used
  boundary_cost <- boundary_length_km * boundary_penalty
  total_cost    <- pu_cost_total + boundary_cost

  tibble(
    method             = label,
    n_selected         = n_selected,
    total_area_km2     = total_area_km2,
    n_patches          = n_patches,
    valid_patches      = valid_patches,
    median_patch_km2   = median_patch_km2,
    pu_cost_total      = pu_cost_total,
    pu_cost_mean       = pu_cost_mean,
    boundary_length_km = boundary_length_km,
    boundary_cost      = boundary_cost,
    total_cost         = total_cost,
    solver_time_sec    = solve_time_sec
  )
}

# baseline metrics (no min patch size, no boundary penalty)
baseline_metrics <- summarise_solution(
  label            = "baseline",
  problem          = p,
  solution_sf      = base_solution,
  solve_time_sec   = base_time[["elapsed"]],
  min_patch_size   = NULL,
  boundary_penalty = 0
)

baseline_metrics


```

------------------------------------------------------------------------

## Approach 1: Manual boundary penalties in prioritizr

### Overview

The baseline solution ignores fragmentation when optimising: the solver only minimises total cost while meeting feature targets. A common way to encourage **more compact** solutions is to add a **boundary penalty** during optimisation.

Here we compare two manually chosen boundary penalties:

-   a **low** penalty, which should gently favour more compact patches, and\
-   a **high** penalty, which should strongly discourage ragged, fragmented solutions.

The downside is that this approach is inherently trial-and-error: it is difficult to predict, from the penalty value alone, how large patches will be or how much cost will increase. Higher penalties can also increase solver runtime, sometimes substantially.

### Run and visualise manual boundary penalties

```{r manual-bp-solve, message = FALSE}
# low boundary penalty
low_boundary_penalty = 0.00001
p_bp_low <- p |>
  add_boundary_penalties(penalty = low_boundary_penalty)

bp_low_time <- system.time({
  bp_low_solution <- solve(p_bp_low)
})

# high boundary penalty
high_boundary_penalty = 0.0001
p_bp_high <- p |>
  add_boundary_penalties(penalty = high_boundary_penalty)

bp_high_time <- system.time({
  bp_high_solution <- solve(p_bp_high)
})

bp_low_time
bp_high_time
```

Visusalise the solutions with low and high boundary penalties

```{r manual-bp-plot-low, fig.width=8, fig.height=6}
plot_prioritizr(bp_low_solution)
```

```{r manual-bp-plot-high, fig.width=8, fig.height=6}
plot_prioritizr(bp_high_solution)
```

### Metrics and interpretation

```{r manual-bp-metrics}
bp_low_metrics <- summarise_solution(
  label = "manual_bp_low",
  problem = p_bp_low,
  solution_sf = bp_low_solution,
  solve_time_sec = bp_low_time[["elapsed"]],
  boundary_penalty = low_boundary_penalty

)

bp_high_metrics <- summarise_solution(
  label = "manual_bp_high",
  problem = p_bp_high,
  solution_sf = bp_high_solution,
  solve_time_sec = bp_high_time[["elapsed"]],
  boundary_penalty = high_boundary_penalty
)

manual_bp_metrics <- dplyr::bind_rows(
  baseline_metrics,
  bp_low_metrics,
  bp_high_metrics
)

manual_bp_metrics
```

We can see that as we set a higher manual boundary, the solve time and total cost increases.

## Approach 2: Automatic calibration with Cohon’s method

### Overview

Choosing boundary penalties by hand can be ad hoc. The function `calibrate_cohon_penalty()` in the *prioritizr* package implements a Cohon-style trade-off calibration. It: 1. Solves the problem for a very chape but fragmented solution 2. Solves the problem for a very compact but expensive solution; and 3. Uses these two anchor solutions to estimate a boundary penalty that balances cost and fragmentation.

### Calibrate and solve

```{r cohon-calibration, message = FALSE}
# add a dummy boundary penalty: calibrate_cohon_penalty() will overwrite it
p_bp_cohon <- p |>
  add_boundary_penalties(penalty = 0.0001)

cohon_time <- system.time({
  cohon_penalty <- calibrate_cohon_penalty(
    p_bp_cohon,
    approx  = TRUE,     # approximate mode for speed
    verbose = TRUE
  )
})

cohon_penalty
cohon_time
```

```{r cohon-solve}
p_bp_cohon_final <- p |>
  add_boundary_penalties(penalty = cohon_penalty)

bp_cohon_time <- system.time({
  bp_cohon_solution <- solve(p_bp_cohon_final, force = TRUE)
})

bp_cohon_time
```

### Visual comparison

```{r plot cohon-calibrated, fig.width=8, fig.height=6}
plot_prioritizr(bp_cohon_solution)

```

### Metrics and comparison with manual penalties

```{r cohon-metrics}
cohon_metrics <- summarise_solution(
  label = "cohon_calibrated",
  problem = p_bp_cohon_final,
  solution_sf = bp_cohon_solution,
  solve_time_sec = cohon_time[["elapsed"]] + bp_cohon_time[["elapsed"]],
  min_patch_size   = NULL,
  boundary_penalty = as.numeric(cohon_penalty)  # 👈 use the calibrated value here
)

# bind everything so far for later comparison
fragmentation_metrics <- dplyr::bind_rows(
  baseline_metrics,
  bp_low_metrics,
  bp_high_metrics,
  cohon_metrics
)

fragmentation_metrics
```

## Approach 3: MinPatch – minimum patch size only

### Overview

Boundary penalties influence fragmentation **during** optimisation. In contrast, the *minpatch* package reduces fragmentation **after** optimisation by enforcing a **minimum patch size** on an existing solution.

Here we start from the **baseline solution** (no boundary penalty) and apply *minpatch* so that every patch is at least a given size. This approach is useful when you already have a rough idea of how large you want each patch to be.

### Set MinPatch parameters

```{r minpatch-params}
# median planning-unit area
median_area <- stats::median(sf::st_area(tas))

# set minimum patch size to 5x the median planning-unit area
min_patch_size <- median_area * 5

# set patch radius to roughly encompass 10 planning units
patch_radius <- sqrt(as.numeric(median_area) * 10)

cat("MinPatch parameters:\n")
cat("- Minimum patch size:", round(min_patch_size, 3), "square meters\n")
cat("- Patch radius:", round(patch_radius, 3), "meters\n")
cat("- This means patches must be at least",
    round(min_patch_size / median_area, 3),
    "times the median planning unit size\n")
```

### Run MinPatch on the baseline solution

```{r minpatch-baseline, message = FALSE}
# time MinPatch processing
minpatch_time_3 <- system.time({
  mp3_result <- run_minpatch(
    prioritizr_problem   = p,
    prioritizr_solution  = base_solution,             # baseline prioritizr solution
    min_patch_size       = min_patch_size,
    patch_radius         = patch_radius,
    boundary_penalty     = 0,             # no boundary penalty inside MinPatch
    remove_small_patches = TRUE,
    add_patches          = TRUE,
    whittle_patches      = TRUE,
    verbose              = TRUE
  )
})

minpatch_time_3
```

### Visualise and summarise MinPatch (patch size only)

```{r plot Min-patch, fig.width=8, fig.height=6}
plot_minpatch(mp3_result, title = "MinPatch Results")
mp3_solution <- mp3_result$solution
mp3_solution <- mp3_solution |>
  dplyr::rename(solution_1 = minpatch)

plot_prioritizr(mp3_solution)
```

```{r minpatch-metrics-3}
mp3_metrics <- summarise_solution(
  label            = "minpatch_min_size",
  problem          = p,
  solution_sf      = mp3_solution,          # output from minpatch
  solve_time_sec   = minpatch_time_3[["elapsed"]],
  min_patch_size   = min_patch_size,           
  boundary_penalty = 0                         # no boundary penalty used
)

mp3_metrics

fragmentation_metrics <- dplyr::bind_rows(
  fragmentation_metrics, # change to fragmentation_metrics once Cohon works
  mp3_metrics
)

fragmentation_metrics
```

We can see from the table that the runtime and the total cost is significantly higher than the manual boundary approach. However, the number of contiguous patches is less, making the solution more compact.

## Approach 4: MinPatch – minimum patch size with boundary penalty

### Overview

MinPatch can also use a **boundary penalty** in its final whittling stage. This does not change the minimum patch size requirement, but it can influence **which edge units are removed or retained**, potentially leading to smoother patch shapes.

Here we repeat Approach 3, but add a small positive boundary penalty inside MinPatch.

### Run and visualise MinPatch with boundary penalty

```{r minpatch-baseline-4, message = FALSE}
minpatch_time_4 <- system.time({
  mp4_result <- run_minpatch(
    prioritizr_problem   = p,
    prioritizr_solution  = base_solution,           # same baseline prioritizr solution
    min_patch_size       = min_patch_size,
    patch_radius         = patch_radius,
    boundary_penalty     = 0.0001,       # small positive penalty for ragged edges
    remove_small_patches = TRUE,
    add_patches          = TRUE,
    whittle_patches      = TRUE,
    verbose              = TRUE
  )
})

minpatch_time_4

# (so summarise_solution() can reuse its logic)
mp4_solution <- tas |>
  dplyr::mutate(solution_1 = mp4_result$solution$minpatch)

```

```{r minpatch-map-compare, fig.width=8, fig.height=6}
plot_minpatch(mp4_result, title = "MinPatch Results")

mp4_solution <- mp4_result$solution
mp4_solution <- mp4_solution |>
  dplyr::rename(solution_1 = minpatch)

plot_prioritizr(mp4_solution)
```

### Metrics and adding to the summary table

```{r minpatch-metrics-4}
mp4_metrics <- summarise_solution(
  label = "minpatch_min_size_bp",
  problem = p,
  solution_sf = mp4_solution,
  solve_time_sec = minpatch_time_4[["elapsed"]],
  min_patch_size = min_patch_size,
  boundary_penalty = 1
)

mp4_metrics

fragmentation_metrics <- dplyr::bind_rows(
  fragmentation_metrics,
  mp4_metrics
)

fragmentation_metrics
```

------------------------------------------------------------------------

## Summary comparison

We can now compare all approaches side by side using the metrics we defined earlier: total cost, total boundary length, number and size of patches, number of features, targets met, and runtime.

```{r summary-table}
fragmentation_metrics |>
  dplyr::arrange(boundary_length_km) |>
  knitr::kable(
    digits = 2,
    caption = "Summary of fragmentation metrics across all approaches."
  )
```

### Side-by-side map comparison

To compare the spatial patterns directly, we can arrange all solutions in a single panel using **patchwork**:

```{r all-maps-patchwork, fig.width = 15, fig.height = 10}
# make sure each call returns a ggplot object
p_base   <- plot_prioritizr(base_solution)       + ggtitle("Baseline")
p_low    <- plot_prioritizr(bp_low_solution)     + ggtitle("Manual BP - Low")
p_high   <- plot_prioritizr(bp_high_solution)    + ggtitle("Manual BP - High")
p_cohon  <- plot_prioritizr(bp_cohon_solution)   + ggtitle("Cohon-calibrated BP")
p_mp3    <- plot_prioritizr(mp3_solution)   + ggtitle("MinPatch")
p_mp4    <- plot_prioritizr(mp4_solution)   + ggtitle("MinPatch + BP")

# arrange in a 2 x 3 grid
(p_base | p_low | p_high) /
(p_cohon | p_mp3 | p_mp4)

```

### Cost–fragmentation trade-off

In this plot, points further left have more compact solutions (short boundaries), and points further down are cheaper. The different approaches occupy different parts of this trade-off space, which helps to clarify when manual penalties, automatic calibration, or MinPatch post-processing might be most useful in a practical workflow.

```{r final-tradeoff-plot, fig.height = 6, fig.width = 8}
ggplot(fragmentation_metrics,
       aes(x = boundary_length_km, y = total_cost,
           colour = method, label = method)) +
  geom_point() +
  ggrepel::geom_text_repel(show.legend = FALSE) +
  theme_minimal() +
  labs(
    x = "Total boundary length (km)",
    y = "Total cost (PU cost + boundary cost)",
    title = "Cost–fragmentation trade-off across all fragmentation controls"
  )

```

### Table summary

| Approach | Method | Pros | Cons |
|------------------|-----------------------------------|------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------------|
| 1 | Manual boundary penalties (*prioritizr*) | Simple; low runtime; easy to implement | Trial-and-error; arbitrary penalty choice; solution patchiness highly sensitive to penalty |
| 2 | Calibrated boundary penalties (`calibrate_cohon_penalty()`) | Systematic trade-off between cost and compactness; reduced subjectivity | High runtime; computationally expensive |
| 3 | Minimum patch size rules (*minpatch*) | Explicit control over patch size; more contiguous solutions; moderate runtime | Depends on chosen minimum patch size; may be less cost-optimal |
| 4 | Boundary penalties within MinPatch | Possibly more precise than Approach 3 | Little to no effect; solutions similar to Approach 3 |
