---
title: "Targets with Mixed Binary + Continuous Features in prioritizr"
author: "Kris Esturas"
date: "2026-03-04"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

library(sf)
library(terra)
library(dplyr)
library(magrittr)
library(prioritizr)
library(exactextractr)
library(spatialplanr) 

set.seed(1)

# Cache downloads here
build_dir <- file.path(getwd(), "_build")
dir.create(build_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Introduction

In spatial conservation planning, we often represent features as binary layers (present/absent) by applying thresholds to continuous data. This is simple, but it can remove nuance — especially when the underlying data contain information about degree of coverage, likelihood of occurrence, or condition/health.

So the practical question is: **how do we include binary and continuous information** in the same prioritisation problem, and set targets accordingly?

This vignette builds the logic through three common questions:

1.   Previously we have considered % coverage of a PU by a feature, but is 2 PUs of 0.25 coverage the same value/importance as 1 PU of 0.5 coverage?

2.   For species/habitat distribution models, it is a probability of occurrence? Do 4 PUs of 20% chance of occurrence equal 1 PU of 80% occurrence?

3.   In the Kosrae work, there is a layer of reef health and we want to prioritise areas of higher reef health. Is this best done with a binary layer and a penalty?

**Key idea used throughout:** targets in *prioritizr* operate on **amounts** that can be summed across PUs. So the practical question becomes:

> How do we turn coverage, probabilities, and condition scores into PU-level amounts that are meaningful to add up — so we can set targets (or objectives) without losing interpretation?

## 2.1 Philippines EEZ as the planning region

We start by using `get_boundary()` from `oceandatr` to obtain the **Philippines EEZ**. This defines our marine planning domain.

```{r planning-domain, echo=TRUE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE, out.width="100%"}
Philippines_eez <- get_boundary(name = "Philippines")

# plot to check we have Philippines' EEZ
p_eez <- splnr_plot(df = Philippines_eez, plotTitle = "EEZ (Philippines)") +
  splnr_gg_add(overlay = Philippines_eez,
               colorOverlay = "black",
               ggtheme = "Default") +
  ggplot2::geom_sf(data = Philippines_eez, fill = "lightgreen", colour = NA)
p_eez
```

## 2.2 Choose and equal-area projection

We reproject the planning region into an equal-area projection to ensure that area- and distance-based calculations are meaningful. For the Philippines, a good default is a *Lambert azimuthal equal-area (LAEA)* projection centred near the country.

```{r projection, echo=TRUE}
# Equal-area projection (meters) centred on the Philippines
projection_Philippines <- "+proj=laea +lon_0=122 +lat_0=12 +datum=WGS84 +units=m +no_defs"
```

## 2.3 Create a planning-unit grid

We now create a grid over the EEZ. To keep runtime moderate, we use a fairly coarse resolution. Our units will be in meters.

```{r PH-grid, echo=TRUE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE, out.width="100%"}
# Grid the planning area
Philippines_grid <- get_grid(
  boundary   = Philippines_eez,
  resolution = 20000, # 30,000 just to test the code but a finer resolution can be opted with a more powerful PC    
  crs        = projection_Philippines,
  output = "sf_hex")


# Plot grid
grid_Philippines <- splnr_plot(df = Philippines_grid, 
                       plotTitle = "Planning grid (Philippines)") +
    splnr_gg_add(Bndry = Philippines_grid,
                 colorBndry = "black",
                 ggtheme = "Default")
grid_Philippines
```

```{r fig.height=12, fig.width=8, message=FALSE, warning=FALSE, out.width="100%"}
reef_zip_url <- "https://wcmc.io/WCMC_008"
reef_zip <- file.path("_build", "WCMC_008_coral_reefs.zip")
reef_dir <- file.path("_build", "WCMC_008_coral_reefs")

# Download
if (!file.exists(reef_zip)) {
  download.file(reef_zip_url, reef_zip, mode = "wb", quiet = TRUE)
}

# Unzip
if (!dir.exists(reef_dir)) {
  dir.create(reef_dir, showWarnings = FALSE)
  unzip(reef_zip, exdir = reef_dir)
}

# Find shapefiles and choose the largest (usually the real reef layer)
shp_files <- list.files(reef_dir, pattern = "\\.shp$", full.names = TRUE, recursive = TRUE)
stopifnot(length(shp_files) >= 1)

reef_shp <- shp_files[which.max(file.info(shp_files)$size)]
cat("Using shapefile:\n", reef_shp, "\n")

reefs_all <- st_read(reef_shp, quiet = TRUE) %>%
  st_make_valid()
```

```{r}
# Ensure EEZ is valid + same CRS as reefs (use EEZ CRS as the reference)
eez <- Philippines_eez %>%
  st_make_valid()

reefs_all <- reefs_all %>%
  st_transform(st_crs(eez))

# 1) Crop by bbox (fast)
reefs_crop <- reefs_all %>%
  st_crop(st_bbox(eez))

# 2) Filter by intersects (keeps features; doesn't cut them yet)
reefs_hit <- reefs_crop %>%
  st_filter(eez, .predicate = st_intersects)

cat("Reefs after crop:", nrow(reefs_crop), "\n")
cat("Reefs intersecting EEZ:", nrow(reefs_hit), "\n")

# 3) Optional: actually cut them to the EEZ boundary
reef_sf <- reefs_hit %>%
  st_intersection(eez)

cat("Reefs after intersection (cut):", nrow(reef_sf), "\n")
```

```{r}
plot(st_geometry(eez), border = "black", lwd = 2)

plot(
  st_geometry(reefs_hit),
  add = TRUE,
  col = grDevices::adjustcolor("yellow", alpha.f = 0.6),
  border = grDevices::adjustcolor("orange", alpha.f = 0.8),
  lwd = 0.4
)
```
